Big Ideas in Sports Analytics and Statistical Tools for their Investigation

### 2022-11-07 response_to_reviewers.Rmd

> While Sections 2 and 3 are well done and complete, Sections 4 and 5 are weak and seem incomplete and unpolished.

>Likewise the Opportunities section should be expanded to include the proliferation of Sports Analytics conferences.

> My largest concern is that a cursory discussion of player contribution or player ratings is given and this seems a major oversight of one of the "big ideas" in sports analytics given the focus on this topic in many sports.  See, for example, Mark Glickman's recent presentation on this at the 2022 Joint Statistical Meetings.  I don't think that Hvattum (2019) is sufficient to cover the area of player contribution.

We thank the reviewer for this comment. We have addressed this important concern in multiple ways:

- We do not consider individual player ratings to be within the scope of this paper, because they are covered by Albert (2015) in this same journal, and because their inclusion would double the length of this paper. We have added the following clarification in scope to the Introduction:

"Notably, we omit a treatment of individual player ratings for team sports, since this concept has been covered ably in these pages by Albert (2015), and its inclusion would double the length of this manuscript. We do, however, discuss individual player ratings in the context of one-person teams (e.g., chess, tennis) in Section 4."

- We have added the following sentence to our Section on team strength: "The methods in this Section apply equally well to multiplayer teams and single-player teams."

- We have added to the existing clarification in scope that we are not considering individual player ratings in team sports in the Conclusion by citing a recent paper from Glickman on this subject. "Recent work by Che & Glickman (2022) also addresses this question across sports."

We hope that these changes will satisfy the reviewer's largest concern.

### 2022-11-07 sports_analytics.Rmd
close #40

Doing the work of sports analytics requires computing with data. 
While the sources of sports data are too numerous to list, in Section \ref{sec:tools} we highlight a few computational tools that make this kind of work possible. 
Section \ref{sec:opportunities} lists several opportunities for disseminating work publicly. 
We conclude in Section \ref{sec:conclusion} with a short discussion of some ideas that are not explored in this paper.

@@

A third crucial idea in sports analytics is the estimation of team strength. 
First, we briefly introduce a simple method for estimating team strength based on won-loss record. 
Next, we detail three other more sophisticated methods for estimating team strength in sports through pairwise evaluations.

@@

The first method for estimating team strength to become popular in sports analytics in a measure of expected winning percentage---often called Pythagorean expectation---developed by @james2003new. 
Later, @miller2007derivation derived the formula as a consequence of assuming that runs are generated by two independent Weibull processes.

@@

Subsequent analysts have tried to find the optimal value of $\alpha$ for various time periods. 
This can be done with a few lines of code, after observing that 
$$
  \frac{X^\alpha}{X^\alpha + Y^\alpha} = \frac{1}{1 + (Y/X)^2}
$$
and fitting a non-linear model (see similar discussion in @baumer2021mdsr). 
Figure \ref{fig:pythag} illustrates the quality of the fit in Major League Baseball.

@@

To obtain win probabilities, as an illustration, for the Federer-Nadal matchup, an estimate for the probability of a Federer victory is:
$$
\hat\Pi_{24} = \frac{exp(\hat\beta_2 - \hat\beta_4)}{1 + exp(\hat\beta_2 - \hat\beta_4)} = \frac{exp(1.136 + 0.062)}{1 + exp(1.136 + 0.062)} = 0.768 \,.
$$

@@

Other fundamental questions are addressed elsewhere. 
How significant is the element of chance in a particular sport? 
Given that we know who the best teams are, who are the best players and how can we quantify their relative contributions? 
What combinations of players work best together in a particular sport? In particular, see @Lopez2018 for estimations of the element of chance across four major sports. 
The second question is often addressed using a formulation of *wins above replacement* (WAR)---see @baumer2015openwar and @Yurko2019nflwar for details in baseball and American football. 
The third question is most compelling in sports like basketball, ice hockey, and soccer, where substitutions are common and it is obvious that different combinations of players with different sets of skills will result in squad of varying strengths and weaknesses.
The concept of *plus-minus*, and then *adjusted* plus-minus is frequently applied to address this question (see @hvattum2019comprehensive for a comprehensive overview of applications).

>>>>

Doing the work of sports analytics requires computing with data. 
While the sources of sports data are too numerous to list, in Section \ref{sec:tools} we highlight a few computational tools that make this kind of work possible. 
Section \ref{sec:opportunities} lists several opportunities for disseminating work publicly. 
We conclude in Section \ref{sec:conclusion} with a short discussion of some ideas that are not explored in this paper. 
Notably, we omit a treatment of individual player ratings for team sports, since this concept has been covered ably in these pages by @Albert2015, and its inclusion would double the length of this manuscript.
We do, however, discuss individual player ratings in the context of one-person teams (e.g., chess, tennis) in Section \ref{sec:strength}.

@@

A third crucial idea in sports analytics is the estimation of team strength. 
First, we briefly introduce a simple method for estimating team strength based on won-loss record. 
Next, we detail three other more sophisticated methods for estimating team strength in sports through pairwise evaluations. 
The methods in this Section apply equally well to multiplayer teams and single-player teams.

@@

The first method for estimating team strength to become popular in sports analytics is a measure of expected winning percentage---often called Pythagorean expectation---developed by @james2003new. 
Later, @miller2007derivation derived the formula as a consequence of assuming that runs (in baseball) are generated by two independent Weibull processes.

@@

Subsequent analysts have tried to find the optimal value of $\alpha$ for various time periods. 
This can be done with a few lines of code, after observing that 
$$
  \frac{X^\alpha}{X^\alpha + Y^\alpha} = \frac{1}{1 + (Y/X)^2}
$$
and fitting a non-linear model (see similar discussion in @baumer2021mdsr). 
Figure \ref{fig:pythag} illustrates the quality of the fit in Major League Baseball since 1954.

@@

To obtain win probabilities, as an illustration, for the Federer-Nadal matchup, an estimate for the probability of a Federer victory is:
$$
\hat\Pi_{24} = \frac{\exp(\hat\beta_2 - \hat\beta_4)}{1 + \exp(\hat\beta_2 - \hat\beta_4)} = \frac{\exp(1.136 + 0.062)}{1 + \exp(1.136 + 0.062)} = 0.768 \,.
$$

@@

Other fundamental questions are addressed elsewhere. 
How significant is the element of chance in a particular sport? 
Given that we know who the best teams are, who are the best players and how can we quantify their relative contributions? 
What combinations of players work best together in a particular sport? In particular, see @Lopez2018 for estimations of the element of chance across four major sports. 
The second question is often addressed using a formulation of *wins above replacement* (WAR)---see @baumer2015openwar and @Yurko2019nflwar for details in baseball and American football. 
Recent work by @che2022athlete also addresses this question across sports. 
The third question is most compelling in sports like basketball, ice hockey, and soccer, where substitutions are common and it is obvious that different combinations of players with different sets of skills will result in squad of varying strengths and weaknesses.
The concept of *plus-minus*, and then *adjusted* plus-minus is frequently applied to address this question (see @hvattum2019comprehensive for a comprehensive overview of applications).
